---
title: "Tres Rios AGPP Data Processing"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

This script is used to process Tres Rios aboveground primary productivity field data. This includes the following key functions: 
1)summing plant-specific biomass calculations
2)generating a QA report for biomass calculations which can be used to ID issues in raw plant data
3)calculating relative abundance
4)aggregating biomass by area and by month.


```{r HOUSKEEPING}

#libraries ----
library('plyr') # always load plyr before dplyr
library("dplyr")
library("tidyr")
library("stringr")
library("lubridate")

#import and clean agpp data
agpp_data <- read.csv('/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/data/tr_agpp_data.csv', stringsAsFactors = FALSE)
agpp_data[agpp_data == ''] <- NA # change all missing values to NA
agpp_data <- agpp_data %>%
  filter(transect != "") %>% # remove all the empty rows 
  select(-leaf_22) #deletes the leaf_22 column
agpp_data[5:30] <- sapply(agpp_data[5:30], function(x) as.numeric(as.character(x))) #force all values columns to numeric format (from various)

#import base spatial data and store in variables (site area, etc.)
tr_totalarea = 210336.28/10 #not sure why this number is divided by 10 in the original quadrat total spreadsheets...perhaps to represent total marsh area represented by each transect?
tr_totalarea_quads = tr_totalarea/5 #divides total area by 5 to approximate the amount of area represented by each transect
tr_water_depth_m = 0.263
tr_water_volume = 210336.28*tr_water_depth_m

#converting various date and number formats
agpp_data <- agpp_data %>%
  mutate(date = mdy(date)) %>% #converts to date format using lubridate
  mutate(plant_ID = as.integer(rownames(agpp_data))) %>% #adds an plant_ID (essentially observation #) for each plant
  mutate(num_seed_stems = ifelse(num_seed_stems == 0, NA, num_seed_stems)) #converts values for num_seed_stesm to NA if 0

#define a function leaf_max to calculate the maximum leaf size
#we define our own function here to overcome challenges of using the existing apply and max fucntions - when a planet has no leaves all inputs to max are NA and the -Inf error occurs
leaf_max <- function(x) ifelse(!all(is.na(x)), max(x, na.rm=TRUE), NA)

```

```{r BIOMASS MODELS}
#define each biomass model as a function to be called downstream when biomass is being calculated
#make edits to functions here and they will automatically be applied downstream
#biomass models derived from '~/WEEL/Tres Rios/datasets/tr-aboveground-primary-productivity/models/plant biomass allometric models.xls'

typha_model <- function(x) 
  (0.093755*agpp_results$sum_leaf_lengths) + ((-7.022353)*agpp_results$num_leaves) + ((-0.301245)*agpp_results$longest_leaf) + 33.036984

typha_seeds_model <- function(x)
  (19.263017*agpp_results$cdb) + (0.3038842*agpp_results$stem_height) + (0.9302222*agpp_results$pistillate_length) + (17.475608*agpp_results$pistillate_width) - 98.18849

sam_model <- function(x)
  (3.5525089*agpp_results$cdb) + (0.0156807*agpp_results$stem_height) - 2.29794

stab_sac_scal_model <- function(x)
  (0.070105*agpp_results$stem_height) - 4.59057

stab_sac_seeds_model <- function(x)
  (0.0385071*agpp_results$stem_height) + (0.322009*agpp_results$stem_volume)

scal_seeds_model <- function(x)
  (0.244936*agpp_results$stem_height) + (0.057674*agpp_results$stem_volume)

smar_model <- function(x)
  (2.2194433*agpp_results$cdb) + (0.290014*agpp_results$stem_height) + (0.826226*agpp_results$num_seed_stems) - 1.947697

```

```{r BIOMASS CALCULATIONS}
#this chunk calculates some summary variables and then outputs a dataframe with individual plant biomass values

#calculate summary variables from raw plant data
agpp_results <- agpp_data %>%
  mutate(sum_leaf_lengths = rowSums(agpp_data[,8:29], na.rm = TRUE)) %>%
  mutate(num_leaves = apply(agpp_data[8:29], 1, function(x) length(unique(x)))) %>%
  mutate(longest_leaf = apply(agpp_data[8:29], 1, function(x) leaf_max(x))) %>%
  mutate(stem_area = 3.14159*((cdb/2)^2)) %>% #cm2
  mutate(stem_volume = ((1/3)*(3.14159)*((cdb/2)^2)*stem_height)) #cm3

#calculate biomass for individual plants
#uses a series of ifelse statements that specify which biomass function to call for a given set of plant conditions (species = ___ and seed/stem/pistillate prescence)
agpp_results$biomass <- ifelse(agpp_results$species == "s_americanus", sam_model(), NA)
agpp_results$biomass <- ifelse(agpp_results$species == "s_californicus" & agpp_results$num_seed_stems > 0, scal_seeds_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_californicus" & is.na(agpp_results$num_seed_stems) == TRUE, stab_sac_scal_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_acutus" & is.na(agpp_results$num_seed_stems) == TRUE, stab_sac_scal_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_tabernaemontani" & is.na(agpp_results$num_seed_stems) == TRUE, stab_sac_scal_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_maritimus", smar_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_acutus" & is.na(agpp_results$num_seed_stems) == FALSE, stab_sac_seeds_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "s_tabernaemontani" & is.na(agpp_results$num_seed_stems) == FALSE, stab_sac_seeds_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "t_latifolia" | agpp_results$species == "t_domingensis" & agpp_results$stem_height > 0, typha_seeds_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "t_latifolia" & is.na(agpp_results$leaf_1) == FALSE, typha_model(), agpp_results$biomass)
agpp_results$biomass <- ifelse(agpp_results$species == "t_domingensis" & is.na(agpp_results$leaf_1) == FALSE, typha_model(), agpp_results$biomass)

#troubleshooting
#produces a new data frame biomass_qa to identify any NA biomass values generated by the previous bit of code. can use this to dig deeper into potential QA issues in original plant data (missing or misplaced values, etc.)
agpp_biomass_qa <- filter(agpp_results, is.na(biomass)) #create a filtered spreadsheet to ID any NA's in biomass

#QA
#remove NA or negative values from biomass calculations column for the sake of facilitating downstream calculations
#this should not be relied on - NA issues in the QA sheet should be investigated and resolved
agpp_results$biomass <- ifelse(is.na(agpp_results$biomass) == TRUE | agpp_results$biomass < 0, 0, agpp_results$biomass) #turns NA's in biomass to 0 for the purposes of downstream calculations

#tidying up - removes all raw and summary data and leaves only metadata and plant biomass calculations
agpp_results <- agpp_results %>%
  select(date:species, biomass, stem_area, stem_volume) %>%
  mutate(biomass = as.numeric(biomass)) #converts biomass to numeric




```

```{r BIOMASS SUMS}
#this chunk scales plant biomass values in space to the quadrat level, and then to the whole-system level

#group agpp_results calculations by date, transect, quadrat, species) and create a data frame to store biomass sums by quadrat
agpp_results <- group_by(agpp_results, date, transect, quadrat, species)
quadrat_totals_agpp <- summarize(agpp_results,
                           total_measured = sum(biomass), #sums the total biomass actually measured
                           quadrat_spp_total = total_measured*4) #multiply measured biomass by 4 (quadrat = 0.25 square meters), this puts units in gdw/m2
agpp_results <- ungroup(agpp_results) #remove groupings that were previously applied to this data frame

#spreading quadrat sums out by species
#we want the data arranged so that biomass for each species is aggregated only according to quadrat. as of previous step, species is still a sorting variable
#this also forces the data frame to display results for all species for all quadrats (even if there is no measured biomass for a species)
#from here on out biomass measurements represent values in gdw/m2 (total actually measured in the field * 4)
quadrat_totals_agpp <- quadrat_totals_agpp %>%
  spread(species, quadrat_spp_total) %>%
  group_by(date, transect, quadrat) %>%
  mutate(total_measured = NULL) #deletes total_measured - no longer necessary
quadrat_totals_agpp[, 4:10][is.na(quadrat_totals_agpp[, 4:10])] <- 0 #replaces all NA values with 0
quadrat_totals_agpp <- quadrat_totals_agpp %>%
  summarize(
    s_acutus = sum(s_acutus),
    s_americanus = sum(s_americanus),
    s_tabernaemontani = sum(s_tabernaemontani),
    s_californicus = sum(s_californicus),
    s_maritimus = sum(s_maritimus),
    t_latifolia = sum(t_latifolia),
    t_domingensis = sum(t_domingensis)) %>%
  ungroup() %>%
  mutate(
    total_measured = (s_acutus + s_americanus + s_tabernaemontani + s_californicus + s_maritimus + t_latifolia + t_domingensis), #calculate all biomass for a given quadrat 
    total_biomass = (total_measured*tr_totalarea_quads)/1000) #scale measured biomass to the amount of area represented by each quadrat, and use /1000 to convert g to kg
```

```{r STEM AREA SUMS}
#this chunk scales plant steam area values in space to the quadrat level, and then to the whole-system level

#group agpp_results calculations by date, transect, quadrat, species) and create a data frame to store biomass sums by quadrat
agpp_results <- group_by(agpp_results, date, transect, quadrat, species)
quadrat_totals_stemarea <- summarize(agpp_results,
                           total_area_measured = sum(stem_area), #sums the total stem_volume actually measured
                           quadrat_spp_total = total_area_measured*4) #multiply measured biomass by 4 (quadrat = 0.25 square meters), this puts units in cm2/m2
agpp_results <- ungroup(agpp_results) #remove groupings that were previously applied to this data frame

#spreading quadrat sums out by species
#we want the data arranged so that data for each species is aggregated only according to quadrat. as of previous step, species is still a sorting variable
#this also forces the data frame to display results for all species for all quadrats (even if there is no measured data for a species)
#from here on out measurements represent values in cm2/m2 (total actually measured in the field * 4)
quadrat_totals_stemarea <- quadrat_totals_stemarea %>%
  spread(species, quadrat_spp_total) %>%
  group_by(date, transect, quadrat) %>%
  mutate(total_area_measured = NULL) #deletes total_volume_measured - no longer necessary
quadrat_totals_stemarea[, 4:10][is.na(quadrat_totals_stemarea[, 4:10])] <- 0 #replaces all NA values with 0
quadrat_totals_stemarea <- quadrat_totals_stemarea %>%
  summarize(
    s_acutus = sum(s_acutus),
    s_americanus = sum(s_americanus),
    s_tabernaemontani = sum(s_tabernaemontani),
    s_californicus = sum(s_californicus),
    s_maritimus = sum(s_maritimus),
    t_latifolia = sum(t_latifolia),
    t_domingensis = sum(t_domingensis)) %>%
  ungroup() %>%
  mutate(
    total_area_measured = (s_acutus + s_americanus + s_tabernaemontani + s_californicus + s_maritimus + t_latifolia + t_domingensis), #calculate all stem volume for a given quadrat 
    total_stem_area_cm2 = (total_area_measured*tr_totalarea_quads)) #scale measured stem areas to the amount of area represented by each quadrat
```

```{r RELATIVE ABUNDANCE}
#this chunk calculates relative abundance at the quadrat level

#create a variable to calculate and store relative abundance by quadrat
relative_abundance <- quadrat_totals_agpp %>%
  mutate(s_acutus = s_acutus/total_measured, #divide each spp biomass by the total for that quadrat, then replace the original biomass values with relative abundance
         s_americanus = s_americanus/total_measured,
         s_tabernaemontani = s_tabernaemontani/total_measured,
         s_californicus = s_californicus/total_measured,
         s_maritimus = s_maritimus/total_measured,
         t_latifolia = t_latifolia/total_measured,
         t_domingensis = t_domingensis/total_measured)
relative_abundance[, 4:10][is.na(relative_abundance[, 4:10])] <- 0 #replaces all NaN values with 0 - using is.na because is.nan does not run on data frames

#create a variable to calculate and store average relative abundance by month
monthly_relative_abundance <- relative_abundance %>%
  mutate(month = month(date)) %>%
  mutate(year = year(date)) %>%
  group_by(month, year) %>%
  summarize(
    s_acutus = mean(s_acutus),
    s_americanus = mean(s_americanus),
    s_tabernaemontani = mean(s_tabernaemontani),
    s_californicus = mean(s_californicus),
    s_maritimus = mean(s_maritimus),
    t_latifolia = mean(t_latifolia),
    t_domingensis = mean(t_domingensis)) %>%
  arrange(year, month)
```

```{r MONTHLY TOTALS}

#create monthly total agpp variable
monthly_agpp_totals <- quadrat_totals_agpp %>% #create monthly total spreadsheet and extract columns of interest, then arrange by ascending year order
  mutate(month = month(date)) %>%
  mutate(year = year(date)) %>%
  select(date, month, year, total_biomass) %>% #selects only the columns we are interested in
  group_by(month, year) %>%
  summarize(
    monthly_biomass = sum(total_biomass))

#create monthly total stem data data frame (area + calculate volume)
monthly_stemarea_totals <- quadrat_totals_stemarea %>% #create monthly total spreadsheet and extract columns of interest, then arrange by ascending year order
  mutate(month = month(date)) %>%
  mutate(year = year(date)) %>%
  select(date, month, year, total_stem_area_cm2) %>% #selects only the columns we are interested in
  group_by(month, year) %>%
  summarize(
    monthly_stem_area_cm2 = sum(total_stem_area_cm2)) %>%
  mutate(
    monthly_stem_area_m2 = monthly_stem_area_cm2/10000, # /10000 to convert cm2 to m2
    monthly_underwater_stem_volume_m3 = monthly_stem_area_m2*tr_water_depth_m
  )

#multiply stem areas by water depth (~25cm average) to get total water volume displaced, then divide by 1000000 to get cm3 to m3
  
#arrange monthly totals by year and month, ascending  
monthly_agpp_totals <- arrange(monthly_agpp_totals, year, month) 
monthly_stemarea_totals <- arrange(monthly_stemarea_totals, year, month) 

#add in new formatted date colummn to make life easier when graphing downstream
monthly_agpp_totals <- monthly_agpp_totals %>%
  mutate(date = paste("01", month, year, sep = "/")) %>%
  mutate(date = dmy(date))

#ungroup data
monthly_agpp_totals <- ungroup(monthly_agpp_totals)

#bring in monthly relative abundance to monthly totals
monthly_agpp_totals$s_acutus <- monthly_relative_abundance$s_acutus
monthly_agpp_totals$s_americanus <- monthly_relative_abundance$s_americanus 
monthly_agpp_totals$s_tabernaemontani <- monthly_relative_abundance$s_tabernaemontani 
monthly_agpp_totals$s_californicus <- monthly_relative_abundance$s_californicus
monthly_agpp_totals$s_maritimus <- monthly_relative_abundance$s_maritimus
monthly_agpp_totals$t_latifolia <- monthly_relative_abundance$t_latifolia
monthly_agpp_totals$t_domingensis <- monthly_relative_abundance$t_domingensis

#calculate monthly total biomass by spp using relative abundance
monthly_agpp_totals <- monthly_agpp_totals %>%
  mutate(
    s_acutus = s_acutus*monthly_biomass,
    s_americanus = s_americanus*monthly_biomass,
    s_tabernaemontani = s_tabernaemontani*monthly_biomass,
    s_californicus = s_californicus*monthly_biomass,
    s_maritimus = s_maritimus*monthly_biomass,
    t_latifolia = t_latifolia*monthly_biomass,
    t_domingensis = t_domingensis*monthly_biomass,
    typha_spp = t_latifolia + t_domingensis,
    sac_stab = s_acutus + s_tabernaemontani
    )

#create a data.frame with monthly totals in metric tons (10^3 kg)
monthly_agpp_totals_MT <- monthly_agpp_totals %>%
  mutate(
    monthly_biomass = monthly_biomass/1000,
    s_acutus = s_acutus/1000,
    s_americanus = s_americanus/1000,
    s_tabernaemontani = s_tabernaemontani/1000,
    s_californicus = s_californicus/1000,
    s_maritimus = s_maritimus/1000,
    t_latifolia = t_latifolia/1000,
    t_domingensis = t_domingensis/1000,
    typha_spp = typha_spp/1000,
    sac_stab = sac_stab/1000
    )

```


```{r}
#PRINTING TO CSV
#write certain datasheets as csv to results directory

#biomass
write.csv(monthly_agpp_totals, file = "/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/results/tr_agpp_monthly_totals.csv")
write.csv(monthly_agpp_totals_MT, file = "/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/results/tr_agpp_monthly_totals_MT.csv")
write.csv(quadrat_totals_agpp, file = "/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/results/tr_agpp_quadrat_totals.csv")
write.csv(monthly_relative_abundance, file = "/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/results/tr_monthly_relative_abundance.csv")

#stem area and volume
write.csv(monthly_stemarea_totals, file = "/Users/chrissanchez/Documents/weel/tres rios/datasets/tr-aboveground-primary-productivity/results/tr_stemdata_monthly_totals.csv")

```

```{r}
#QA

#create and write write file to output all unique values for all columns for QA review

tr_agpp_metadataQA <- lapply(agpp_data, unique)
write.csv(tr_agpp_metadataQA, file = "/Volumes/GoogleDrive/My Drive/weel/tres rios/datasets/tr-aboveground-primary-productivity/data/tr_metadataQA")

```


```{r}
#BIOMASS RATES - may just need to continue doing this in excel for now until I figure out how to do this step in R

#create a new data frame that calculates the differences between each consecutive month (row) in monthly_totals for each plant species
biomass_rates <- data.frame(
  typha_diff = abs(diff(monthly_totals$typha_spp)), 
  sac_stab_diff = abs(diff(monthly_totals$sac_stab)),
  s_americanus_diff = abs(diff(monthly_totals$s_americanus)),
  s_californicus_diff = abs(diff(monthly_totals$s_californicus)),
  s_maritimus_diff = abs(diff(monthly_totals$s_maritimus)))

#introduce a formula to calculate moving averages
mav <- function(x,n=5){filter(x,rep(1/n,n), sides=2)}

  #LEFT OFF HERE - one trick may be using the "lags" function in dplyr to do something?
  # moving averages may also be key to getting the biomass rates to calculate correctly
  # finding a way to autopopulate dates (may need to write custom function) could also help tremendously
  
```


*To Do List*
1) Verify that NA coercion is not creating issues downstream (missing values, etc.).

2) Investigate how to automate QAQC.

3) Re-arrange columns in monthly_totals to put formatted date column first

4) ~~Relative abundance is calculating incorectly - not summing to 100~~ *FIXED 8/5/2017*

5) make sure that field data columns primary productivty data sheet is formatted as numeric in order to write script to calculate plant characteristics
6) write scripts to calculate plant characteristics
6)manually calculate biomass using allometric models
6) investigate QA with regards to thatched quadrates and NAs

20181211
added stuff to calculate stem volumes, need to do the same for stem areas
2) is the "quadrat" label misleading? since values in those spreadsheets are not actually #'s measured, but, multipled by 4 to put it in terms of area?

EXTRA CODE BITS
#below is code scrapped from an attempt to reorgnize how plant data is structured - may be useful later
primary_productivity[,8:29] <- sapply(primary_productivity[,8:29], function(x) as.numeric(as.character(x)))
primary_names <- colnames(primary_productivity)
primary_names <-sub("leaf_", "leaf__", primary_names)
colnames(primary_productivity) <- primary_names


#new approach to primary productivity - link everything to plant_ID 
primary_productivity_new <- primary_productivity %>%
  mutate(plant_ID = as.integer(rownames(primary_productivity))) %>%
  select(1:31, data.book.ID, Notes, plant_ID) %>%
  gather(characteristic, value, cdb:pistillate_width, na.rm = TRUE, convert = TRUE) %>%
  mutate(date = ymd(date)) %>%
  na.omit(primary_productivity_new$value) %>% #remove any NA's from value column to allow us to 
  mutate(value = as.numeric(value)) %>%
  separate(characteristic, into = c("characteristic", "characteristic_type"), sep = "__", fill = "right", convert = TRUE)  %>%
  arrange(plant_ID)

primary_productivity_results <- primary_productivity_new %>%
  group_by(date, transect, quadrat, species, plant_ID, characteristic) %>%
  summarize(
    value = sum(value)
  )

#old nested ifelse statement

agpp_results$biomass <-  ifelse(agpp_results$species == "s_americanus", sam_model(), 
                         ifelse(agpp_results$species == "s_californicus" & agpp_results$num_seed_stems > 0, scal_seeds_model(), 
                         ifelse(agpp_results$species == "s_acutus" | agpp_results$species == "s_californicus" | agpp_results$species == "s_tabernaemontani" & is.na(agpp_results$num_seed_stems) == TRUE,           stab_sac_scal_model(),
                         ifelse(agpp_results$species == "s_maritimus", smar_model(),
                         ifelse(agpp_results$species == "s_acutus" | agpp_results$species == "s_tabernaemontani" & agpp_results$num_seed_stems > 0, stab_sac_seeds_model(),
                         ifelse(agpp_results$species == "t_latifolia" | agpp_results$species == "t_domingensis" & agpp_results$pistillate_length > 0, typha_seeds_model(),
                         ifelse(agpp_results$species == "t_latifolia" | agpp_results$species == "t_domingensis", typha_model(), "leftovers")
                         ))))))
